Field-informed Reinforcement Learning for Learning Large Scale Collective Tasks

### 2023-05-10 paper.tex
some intro, background and BE fix

\begin{IEEEkeywords}
aggregate computing, Graph Neural Network, Cyber-Physical Swarms.
\end{IEEEkeywords}
%
\ga{Paper levels:
\begin{itemize}
  \item Distributed collective intelligence
  \item GNN and field coordination
  \item Spatial tracking
\end{itemize}
}
\section{Introduction}
Various phenomena in the real world are not evenly distributed across the environment such as watershed, wild fires, traffic, human crowd movement, or wild life movement from insects, fish, birds, and mammals.
To acquire appropriate information about such phenomena, we also need to distribute the sensors accordingly to cover all aspects sufficiently. While manual deployment will mitigate the problem, not all phenomena are known exactly \emph{a priori} and would require maintenance and adjustment during runtime. Even worse, with phenomena able to change their location, size, and distribution---as it is the case with crowds, wild life, or wild fires---an adaptation of the collective sensors is required in order to keep the required information.

@@

In this paper, we propose a novel approach combining aggregate computing~\cite{Beal2015Computer} with Graph Neural Networks~\cite{Zhou2020AIOpen} to focus on the relevant areas. We further combine this with reinforcement learning techniques to respond to changing conditions in the environment. 
%
This follow a novel trends in which high-level declarative programming language are mixed with machine learning technique in order to synthesise robust collective controllers~\cite{DBLP:conf/acsos/Aguzzi21,DBLP:conf/icdcs/AguzziCV22,DBLP:conf/coordination/AguzziCV22}.
%Specifically, we will show... \lukas{more info...}
Specifically, 
 our approach utilizes aggregate computing to disseminate information by manipulating a computational field, 
 which is a distributed data structure that maps devices to information. 
 This field functions as a dynamic environment in which information is continuously updated and diffused throughout the system. 
 By leveraging this layer, we can build collective distributed intelligence using \ac{GNN},
  which uses the information in the field to determine the best action for a given task.
\ga{first draft, need a refinement}
The remainder of this paper is structured as follows. First, we introduce the relevant background and problem formulation in \Cref{sec:background}. Afterwards, we introduce our approach in \Cref{sec:approach}. \Cref{sec:eval} outlines the performed experiments and discusses the obtained results. Finally, we will present our conclusions in \Cref{sec:conclusion}.

\section{Background and Motivation}
\label{sec:background}
In this section, we aim to provide a concise overview of the system we intend to address, 
 starting with a system definition. 

@@

We then proceed to discuss relevant research areas that inform our approach. 

@@

Finally, we present a formalization of the problem and a discussion of the motivations behind our proposed work.
\ga{Page budget: 1/2 page \\}
\ga{Plan: discussion about the problem of coordination in multi-agent systems 
and the need for a scalable approach. In doing this, we will discuss some of the existing approaches (declarative, RL and )}
\subsection{System definition}
This article focuses on the definition of \emph{decentralized collective intelligence} within the context of \emph{large-scale} distributed systems, 
 specifically what we call \emph{Cyber-Physical Swarms} or \emph{swarm-like systems}
 that are systems of \emph{devices} (or nodes) equipped with \emph{sensors} and \emph{actuator}
 each of which can only interact with the \emph{1-hop} neighbourhood that can 
 be logical (i.e., a communication network) or physical (i.e., a spatial neighbourhood).

@@

Each node executes a \emph{local} controller that can only access the information available to its \emph{context}
 that is the information available in the neighbourhood and the information available from its sensors.
After each program evaluation, 
 the node can send messages to its neighbourhood about its current evaluation and update its internal state.
This general model can be adapted to several different scenarios, 
 such as a network of sensors, a network of robots (e.g., swarm robotics), or a network of mobile phones.

@@

 such as spatial area coverage, phenomena tracking, and robot aggregation.

@@

\subsection{Field Coordination Approaches}

@@

The field-based coordination approaches utilize a concept of \emph{computational fields} 

@@

A modern field-coordination approach called \emph{aggregate computing}~\cite{Beal2015Computer}
 is rooted in earlier work on \emph{artificial potential fields}~\cite{DBLP:conf/icra/Warren89} and \emph{co-fields}~\cite{DBLP:journals/pervasive/MameiZL04}. 

@@

 In this \emph{macro-programming}~\cite{DBLP:journals/corr/abs-2201-03473} paradigm is possible to devise collective and self-organizing behaviour through a composition of functions operating on fields. 

@@

@@

Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio),
 \ac{GNN} is a novel model used to process graph-structured data. 

@@

and $V$ identifies the nodes present in the graph at that time. 

@@

Each node $v$ is associated with an observation (or feature set), $o_v$, 

@@

Given $o_v$ a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$.

@@

@@

where $h_{v}^{k}$ is the embedding of node $v$ at the $k$-th layer ($h_v^0 = o_v$), 
 $\mathcal{N}(v)$ is the set of neighbors of node $v$ computed from $E$.

@@

The differential part come into play in the $\operatorname{COMBINE}$ function, 

@@

  and it could be a simple sum or a more complex function such as a neural network layer.

@@

\acp{GNN} are used in several fields such as social network analysis, chemistry, and physics.
In this paper, we use GNNs to learn a local behaviour for each agent in a multi-agent system (more details in Section~\ref{sec:approach}).
\subsection{Many-Agent Reinforcement Learning}
\ga{plan: we could discuss briefly about standard approach (few agents), and mean-field approach (many agents). 
This will lead to our approach, which is a combination of DQN (or any Deep RL approach) and GNN.}

\Ac{RL} has gained a lot of interest recently, 
 thanks to its successful application in various scenarios, 
 ranging from video games (such as Alpha Go~\cite{Silver2016Go} and Atari~\cite{Atari2016DQN}) 
 to chatbots (like ChatGPT~\cite{ChatGPT2023}). 
% 
In \ac{RL}, an agent (i.e., a smart entity capable of making decisions) 
 performs actions in an environment (i.e., everything outside the agent) according to a policy, 
 to maximize long-term rewards.

@@

In fact, the first works in this direction were discussed precisely in the last field, 
 exploring new models (e.g., swarMDP~\cite{DBLP:conf/atal/SosicKZK17}) and learning algorithms capable of extrapolating 
 a policy equal to the entire system. 
Modern approaches, however, 
 have started to consider the use of deep learning approaches to try 
 to synthesize robust controllers capable of generalizing to new tasks. 
In this context, mean-field reinforcement learning~\cite{pmlr-v80-yang18d} is certainly noteworthy. 
 Mean-field RL is a technique for learning approximate inference in large-scale systems, 
 where the interactions between the agents can be decoupled into a single conditional probability distribution function for each agent, 
 thus simplifying the learning process and reducing the computational cost.
%
Some known approaches using mean-field reinforcement learning include Q-mean, 
 which is an extension of Q-learning to mean-field settings, 
 and actor-critic mean-field, which combines actor-critic algorithms with mean-field approximations. 
%
These approaches have shown promising results in various domains, such as multi-agent coordination 
 and decentralized control, and are actively being researched and developed for further applications
\subsection{Problem formalisation}
Given the homogeneity and large scale considered and the \emph{locality} (i,e., each agent can only observe its neighbours), 
 the problem can be modelled through the SwarMDP model~\cite{DBLP:conf/atal/SosicKZK17} -- 
 an extension of the \ac{DecPOMDP}~\cite{Decpomdp2000} model to swarm-like systems.
% 
A SwarMDP is characterized by a \emph{swarming agent} ($\mathbb{A}$) and the dynamics of the environment ($\mathbb{E}$).
Specifically, $\mathbb{A}$ is a tuple ($\mathcal{S}, \mathcal{O}, \mathcal{A}, \mathcal{R}, \pi$) where:
\begin{itemize}
  \item $\mathcal{S, O, A}$ are the set of local states, observations (or features), and actions, respectively;
  \item $\mathcal{R}: \mathcal{S} \rightarrow \mathbb{R}$ is the reward function, which is influenced by the environment;
  \item $\pi: \mathcal{O} \rightarrow \mathcal{A}$ is the policy function, which maps the observations to the actions. It could be deterministic or stochastic.
\end{itemize}
Starting from this definition, the environment $\mathbb{E}$ is defined as a tuple ($\mathcal{P}, \mathbb{A}, \mathcal{T}, \xi$), where:
\begin{itemize}
  \item $\mathcal{P}$ is the total number of agents in the systems (the agent population), which is assumed to be fixed;
  \item $\mathbb{A}$ is the above defined agent prototype that rules each agent $v \in P$;
  \item $\mathcal{T}: \mathcal{S}^P \times \mathcal{A}^P \times \mathcal{S}^P \rightarrow \mathbb{R}$ is the transition  global function, which is influenced by the actions of the agents and returns a collective reward -- this is typically not known by the swarming agents;
  \item $\xi: \mathcal{S^P} \rightarrow \mathcal{O^P}$ is the global observation model of the systems.
\end{itemize}
In swarMDP, the neighbourhood is not directly defined, but it is implicitly defined by the observation model $\xi$.
Since in our case, the agents can only interact with 1-hop neighbourhoods and they aren't directly influenced by other 
  agent observations, we restrict the observation model as follows:
\begin{equation}
\xi(v): \{s_j, j \in \mathcal{N}^v\} \rightarrow O
\end{equation}
\begin{equation}
\xi = \{\xi(v), v \in \mathcal{P}\}
\end{equation}
where $\mathcal{N}^v$ is the set of neighbours of $v$.

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\begin{equation}
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\end{equation}
\begin{equation}
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
\end{equation}
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation $o_v$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022}.
\subsection{Motivation}
The definition of distributed controllers for intelligent systems can be divided into two major approaches: 
 \emph{manual design}, where the application is conceived and described through a programming language, 
 and \emph{automatic design}~\cite{brambilla2013swarm}, which employs machine learning techniques to synthesize distributed controllers. 
%
Our work lies between these two approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.

Compared to earlier works on field-based coordination, 
 our approach builds on the concept of co-fields~\cite{DBLP:journals/pervasive/MameiZL04}, 
 where agents exploit the field as a stigmergic medium to receive system-wide information and apply reasoning to this data.
%  
We present a subsequent approach where agent intelligence is synthesized through \ac{MAARL} and the use of \ac{GNN} to learn a local representation from the neighbourhood.
%
By learning a smart policy directly in the environment, the agent becomes capable of adapting its behaviour to new situations.

@@

In this section, we aim to clarify the components involved in our proposed solution (i.e., architecture), how these components interact with each other to bring the system to perform the collective behavior (i.e dynamics), and finally, we will detail the learning algorithm designed and used to synthesize the policy (learning algorithm).

@@

Despite appearing centralized, this system proposes a completely decentralized execution. 

@@

and the \ac{GNN} can be locally evaluated using only neighborhood information. 

@@

We want to emphasize that, in this case, the \acp{GNN} must be 1-hop; 

@@

In our case, we leveraged the property of \ac{GNN}s to have a dual interpretation, i.e., to function globally over the entire graph and locally only over the neighborhood. 
%
Therefore, we relied on the value-based approach \ac{DQN}~\cite{mnih2015playing} \ga{to introduce in many rl??} with two major modifications: 
\begin{enumerate}
  \item experience replay stores experiences in the form of \emph{graphs},
  \item the neural network used to compute the Q function is based on a \ac{GNN} with an \ac{MLP} downstream.
\end{enumerate}
The first point is a natural extension due to the fact that we work on graphs rather than simple values. 
%
For the second point,
 the use of \acp{GNN} allows us to define policies on a variable neighborhood  that is essential in such systems as this can vary based on the neighborhood policy,
 and it is known that \acp{GNN} have a certain ability to generalize to new structures and scale with different nodes. 
% 
Additionally, 
 using the overall graph compared to local experiences makes learning more stable as it reduces the non-stationarity of the environment perceived by each node.
% 
 During learning, it is as if we have a \emph{single} agent collectively choosing which action to take (i.e., resulting action graph). 
%
\ga{todo: describe the algorithm with the modifications}
\begin{algorithm}
  \KwIn{Environment $\mathbb{E}$, graph replay buffer $\mathcal{D}$, target network $\theta^{-}$, current network $\theta$, exploration strategy $\epsilon$}
  \KwOut{Trained DQN model $\theta$}

  Initialize $\mathcal{D}$ with random initial transitions;

  Initialize $\theta$ with random weights;

@@

\section{Evaluation}
\label{sec:eval}
\ga{Page budget: 2 pages \\}
\ga{Plan: we should discuss the robot aggregation scenario, specifying the state space, the action space, and the reward function.
I dunno if it makes sense to discuss the variants (i.e., fixed area, two areas and moving area). We will see.
}
To test the effectiveness of the proposed approach, 
 we decided to experiment with a case study related to swarm robotics, 
 specifically the tracking and coverage of a spatiotemporal phenomenon, cf. tracking a wildfire 
 or monitoring the water levels in a canal.

\subsection{Scenario}
The proposed simulation consists 
 of a set of robots placed in a 2D space capable of perceiving the intensity 
 of the phenomenon of interest through an installed sensor $\zeta_v$ with $v \in \mathbb{V}$. 
 (e.g., camera, temperature sensor, etc.). 
%
Additionally, each node has a coverage range $\omega$ 
 that describes the area it can monitor. 
 Each drone can only communicate directly with its own neighborhood $\mathcal{N}$, 
 which in this case depends on a $\mathbb{O}$ range,
From the neighbourhood, 
 each node can also perceive the direction of the other nodes.
%
Each node moves following a certain action composed of two components ($\alpha$, $\theta$) 
 which respectively describe the angle and intensity of the movement (i.e., the velocity vector).
\subsection{Goal}
 The objective of this scenario is to:
\begin{itemize}
\item Maximize the number of drones within the phenomena
\item Minimize the number of drones without a neighbourhood
\item Maximize the coverage of the system
\end{itemize}

@@

 Finally, to maximize the coverage, 
  we decided to calculate the maximum and minimum distance of the neighbourhood and give more value in case these two values are similar. 
%
This means that the average distance will tend towards the one expressed by the viewing range of each node:
% 
\begin{equation*}
R^{C}_{v} = \frac{d_{max} - d_{min}}{\mathbb{O}}
\end{equation*}
The final reward function is defined as:
\begin{equation*}
R_{v} = R^a_{v} + R^{\mathcal{N}}_{v} + R^{C}_{v}
\end{equation*}

>>>>

\begin{IEEEkeywords}
aggregate computing, Graph Neural Network, Cyber-Physical Swarms.
\end{IEEEkeywords}
%
\ga{Paper levels:
\begin{itemize}
  \item Distributed collective intelligence
  \item GNN and field coordination
  \item Spatial tracking
\end{itemize}
}
\section{Introduction}

Autonomous agents interacting and operating towards a common task, rely on shared information and consequently emerging knowledge in order to achieve the underlying goals efficiently. Coordinating and organising this information exchange within the network of agents is required for improve the collaboration among them further~\cite{jennings1996coordination}. Through this coordination, agents may overcome problems requiring (i) efficiency and accuracy only achieved by multiple agents, (ii) the complimentary competences of the individual agents, or (iii) the collective resources of multiple agents.

Various phenomena in the real world are not evenly distributed across the environment such as watershed, wild fires, traffic, human crowd movement, or wild life movement from insects, fish, birds, and mammals.
To acquire appropriate information about such phenomena, we also need to distribute the sensors accordingly to cover all aspects sufficiently. While manual deployment will mitigate the problem, not all phenomena are known exactly \emph{a priori} and would require maintenance and adjustment during runtime. Even worse, with phenomena able to change their location, size, and distribution---as it is the case with crowds, wild life, or wild fires---an adaptation of the collective sensors is required in order to keep the required information.

@@

In this paper, we propose a novel approach combining aggregate computing~\cite{Beal2015Computer} with Graph Neural Networks~\cite{Zhou2020AIOpen} to focus on the relevant areas. We further combine this with reinforcement learning techniques to respond to changing conditions in the environment. 
%
This follow a novel trends in which high-level declarative programming language are mixed with machine learning technique in order to synthesise robust collective controllers~\cite{DBLP:conf/acsos/Aguzzi21,DBLP:conf/icdcs/AguzziCV22,DBLP:conf/coordination/AguzziCV22}.
%Specifically, we will show... \lukas{more info...}
Specifically, 
 our approach utilises aggregate computing to disseminate information by manipulating a computational field, 
 which is a distributed data structure that maps devices to information. 
 This field functions as a dynamic environment in which information is continuously updated and diffused throughout the system. 
 By leveraging this layer, we can build collective distributed intelligence using \ac{GNN},
  which uses the information in the field to determine the best action for a given task.
\ga{first draft, need a refinement}
The remainder of this paper is structured as follows. First, we introduce the relevant background and problem formulation in \Cref{sec:background}. Afterwards, we introduce our approach in \Cref{sec:approach}. \Cref{sec:eval} outlines the performed experiments and discusses the obtained results. Finally, we will present our conclusions in \Cref{sec:conclusion}.

\section{Background and Motivation}
\label{sec:background}
In this section, we aim to provide a concise overview of the system we intend to address, 
 starting with a system definition. 
%
We then proceed to discuss relevant research areas that inform our approach. 
%
Finally, we present a formalization of the problem and a discussion of the motivations behind our proposed work.
\ga{Page budget: 1/2 page \\}
\ga{Plan: discussion about the problem of coordination in multi-agent systems 
and the need for a scalable approach. In doing this, we will discuss some of the existing approaches (declarative, RL and )}
\subsection{System definition}
This article focuses on the definition of \emph{decentralised collective intelligence} within the context of \emph{large-scale} distributed systems.
 Specifically, we focus on \emph{Cyber-Physical Swarms} or \emph{swarm-like systems}, i.e.,  systems consisting of \emph{devices} (or nodes) equipped with \emph{sensors} and \emph{actuator} able to interact with the real world.
 Each of these nodes can interact with its direct neighbours, either based on physical (e.g. communication range) or logical distance (e.g. 1-hop neighbourhood).
 \lukas{does the article focus on the definition of decentralised collective intelligence?}
  \lukas{distributed or decentralised?}

@@

Each node is governed by an autonomous software agent, executing a \emph{local} control loop. In each iteration of this loop, the agent can access the information available within its own \emph{context}. This context is comprised by information acquired by the node directly through its sensors and the information received from its neighbours. 

@@

At the end of each control loop iteration, the node can send messages to its neighbourhood. This message may contain raw sensor data or aggregated information over time or from other neighbours.
% about its current evaluation and update its internal state.

@@

%This general model can be adapted to several different scenarios, such as a network of sensors, a network of robots (e.g., swarm robotics), or a network of mobile phones.
Examples of such large scale system can be a network of robots (e.g., swarm robotics), camera and IoT networks, or a network of mobile phones.

@@

 such as spatial area coverage, phenomena tracking, and node aggregation.

@@

\lukas{not really a system definition - maybe we rename this to \emph{Swarm Systems}?}
\subsection{Field-based Coordination}

@@

The field-based coordination approaches utilise a concept of \emph{computational fields} 

@@

\emph{Aggregate computing}~\cite{Beal2015Computer}, a modern field-coordination approach, is rooted in earlier work on \emph{artificial potential fields}~\cite{DBLP:conf/icra/Warren89} and \emph{co-fields}~\cite{DBLP:journals/pervasive/MameiZL04}. 

@@

Within aggregate computing, the  \emph{macro-programming}~\cite{DBLP:journals/corr/abs-2201-03473} paradigm can be realised to devise collective and self-organizing behaviour through a composition of functions operating on fields. 

@@

\lukas{still not sure how the coordination works or if it has been explained in the above paragraph.}

@@

%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.

@@

and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}

@@

Each node $v \in V$ is associated with an observation (or feature set) $o_v$.

@@

Given $o_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. The node embedding $h_v$ describes the node in the network and summarise the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.

@@

  \lukas{maybe add:``The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. With each traversal of an edge $e \in E$ from $v$, the number of layers increased.''}

@@

where $h_{v}^{k}$ is the embedding of node $v$ within the $k$-th layer ($h_v^0 = o_v$), 
 $\mathcal{N}(v)$ is the set of neighbours of node $v$ computed from $E$.

@@

The differential part comes into play in the $\operatorname{COMBINE}$ function, 

@@

  and it could be a simple sum, max or sum of products. More complex aggregation functions are available~\cite{pellegrini2020learning}.

@@

\acp{GNN} are used in several application areas such as social network analysis, chemistry, and physics.
In this paper, we use GNNs to learn a local behaviour for each agent in a multi-agent system (more details in Section~\ref{sec:approach}).
\subsection{Many-Agent Reinforcement Learning}
\ga{plan: we could discuss briefly about standard approach (few agents), and mean-field approach (many agents). 
This will lead to our approach, which is a combination of DQN (or any Deep RL approach) and GNN.}

\Ac{RL} has gained a lot of interest recently, 
 thanks to its successful application in various scenarios, 
 ranging from video games (such as Alpha Go~\cite{Silver2016Go} and Atari~\cite{Atari2016DQN}) 
 to chatbots (like ChatGPT~\cite{ChatGPT2023}). 
% 
In \ac{RL}, an agent (i.e., a smart entity capable of making decisions) 
 performs actions in an environment (i.e., everything outside the agent) according to a policy, 
 to maximise long-term rewards (or minimise a regret).

@@

In fact, the first works in this direction were discussed precisely in the last field, 
 exploring new models (e.g., swarMDP~\cite{DBLP:conf/atal/SosicKZK17}) and learning algorithms capable of extrapolating 
 a policy equal to the entire system. 
Modern approaches, however, 
 have started to consider the use of deep learning approaches to try 
 to synthesise robust controllers capable of generalizing to new tasks. 
In this context, mean-field reinforcement learning~\cite{pmlr-v80-yang18d} is certainly noteworthy. 
 Mean-field RL is a technique for learning approximate inference in large-scale systems, 
 where the interactions between the agents can be decoupled into a single conditional probability distribution function for each agent, 
 thus simplifying the learning process and reducing the computational cost.
%
Some known approaches using mean-field reinforcement learning include Q-mean, 
 which is an extension of Q-learning to mean-field settings, 
 and actor-critic mean-field, which combines actor-critic algorithms with mean-field approximations. 
%
These approaches have shown promising results in various domains, such as multi-agent coordination 
 and decentralised control, and are actively being researched and developed for further applications
\subsection{Problem formalisation}
Given the homogeneity and large scale considered and the \emph{locality} (i,e., each agent can only observe its neighbours), 
 the problem can be modelled through the SwarMDP model~\cite{DBLP:conf/atal/SosicKZK17} -- 
 an extension of the \ac{DecPOMDP}~\cite{Decpomdp2000} model to swarm-like systems.
% 
A SwarMDP is characterised by a \emph{swarming agent} ($\mathbb{A}$) and the dynamics of the environment ($\mathbb{E}$).
Specifically, $\mathbb{A}$ is a tuple ($\mathcal{S}, \mathcal{O}, \mathcal{A}, \mathcal{R}, \pi$) where:
\begin{itemize}
  \item $\mathcal{S, O, A}$ are the set of local states, observations (or features), and actions, respectively;
  \item $\mathcal{R}: \mathcal{S} \rightarrow \mathbb{R}$ is the reward function, which is influenced by the environment;
  \item $\pi: \mathcal{O} \rightarrow \mathcal{A}$ is the policy function, which maps the observations to the actions. It could be deterministic or stochastic.
\end{itemize}
Starting from this definition, the environment $\mathbb{E}$ is defined as a tuple ($\mathcal{P}, \mathbb{A}, \mathcal{T}, \xi$), where:
\begin{itemize}
  \item $\mathcal{P}$ is the total number of agents in the systems (the agent population), which is assumed to be fixed;
  \item $\mathbb{A}$ is the above defined agent prototype that rules each agent $v \in P$;
  \item $\mathcal{T}: \mathcal{S}^P \times \mathcal{A}^P \times \mathcal{S}^P \rightarrow \mathbb{R}$ is the transition  global function, which is influenced by the actions of the agents and returns a collective reward -- this is typically not known by the swarming agents;
  \item $\xi: \mathcal{S^P} \rightarrow \mathcal{O^P}$ is the global observation model of the systems.
\end{itemize}
In swarMDP, the neighbourhood is not directly defined, but it is implicitly defined by the observation model $\xi$.
Since in our case, the agents can only interact with 1-hop neighbourhoods and they aren't directly influenced by other 
  agent observations, we restrict the observation model as follows:
\begin{equation}
\xi(v): \{s_j, j \in \mathcal{N}^v\} \rightarrow O
\end{equation}
\begin{equation}
\xi = \{\xi(v), v \in \mathcal{P}\}
\end{equation}
where $\mathcal{N}^v$ is the set of neighbours of $v$.

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\begin{equation}
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\end{equation}
\begin{equation}
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
\end{equation}
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation $o_v$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022}.
\subsection{Motivation}
The definition of distributed controllers for intelligent systems can be divided into two major approaches: 
 \emph{manual design}, where the application is conceived and described through a programming language, 
 and \emph{automatic design}~\cite{brambilla2013swarm}, which employs machine learning techniques to synthesise distributed controllers. 
%
Our work lies between these two approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.

Compared to earlier works on field-based coordination, 
 our approach builds on the concept of co-fields~\cite{DBLP:journals/pervasive/MameiZL04}, 
 where agents exploit the field as a stigmergic medium to receive system-wide information and apply reasoning to this data.
%  
We present a subsequent approach where agent intelligence is synthesised through \ac{MAARL} and the use of \ac{GNN} to learn a local representation from the neighbourhood.
%
By learning a smart policy directly in the environment, the agent becomes capable of adapting its behaviour to new situations.

@@

In this section, we aim to clarify the components involved in our proposed solution (i.e., architecture), how these components interact with each other to bring the system to perform the collective behavior (i.e dynamics), and finally, we will detail the learning algorithm designed and used to synthesise the policy (learning algorithm).

@@

Despite appearing centralised, this system proposes a completely decentralised execution. 

@@

and the \ac{GNN} can be locally evaluated using only neighbourhood information. 

@@

We want to emphasise that, in this case, the \acp{GNN} must be 1-hop; 

@@

In our case, we leveraged the property of \ac{GNN}s to have a dual interpretation, i.e., to function globally over the entire graph and locally only over the neighbourhood. 
%
Therefore, we relied on the value-based approach \ac{DQN}~\cite{mnih2015playing} \ga{to introduce in many rl??} with two major modifications: 
\begin{enumerate}
  \item experience replay stores experiences in the form of \emph{graphs},
  \item the neural network used to compute the Q function is based on a \ac{GNN} with an \ac{MLP} downstream.
\end{enumerate}
The first point is a natural extension due to the fact that we work on graphs rather than simple values. 
%
For the second point,
 the use of \acp{GNN} allows us to define policies on a variable neighbourhood  that is essential in such systems as this can vary based on the neighbourhood policy,
 and it is known that \acp{GNN} have a certain ability to generalise to new structures and scale with different nodes. 
% 
Additionally, 
 using the overall graph compared to local experiences makes learning more stable as it reduces the non-stationarity of the environment perceived by each node.
% 
 During learning, it is as if we have a \emph{single} agent collectively choosing which action to take (i.e., resulting action graph). 
%
\ga{todo: describe the algorithm with the modifications}
\begin{algorithm}
  \KwIn{Environment $\mathbb{E}$, graph replay buffer $\mathcal{D}$, target network $\theta^{-}$, current network $\theta$, exploration strategy $\epsilon$}
  \KwOut{Trained DQN model $\theta$}

  Initialise $\mathcal{D}$ with random initial transitions;

  Initialise $\theta$ with random weights;

@@

\section{Evaluation}
\label{sec:eval}
\ga{Page budget: 2 pages \\}
\ga{Plan: we should discuss the robot aggregation scenario, specifying the state space, the action space, and the reward function.
I dunno if it makes sense to discuss the variants (i.e., fixed area, two areas and moving area). We will see.
}
To test the effectiveness of the proposed approach, 
 we decided to experiment with a case study related to swarm robotics, 
 specifically the tracking and coverage of a spatio-temporal phenomenon, cf. tracking a wildfire 
 or monitoring the water levels in a canal.

\subsection{Scenario}
The proposed simulation consists 
 of a set of robots placed in a 2D space capable of perceiving the intensity 
 of the phenomenon of interest through an installed sensor $\zeta_v$ with $v \in \mathbb{V}$. 
 (e.g., camera, temperature sensor, etc.). 
%
Additionally, each node has a coverage range $\omega$ 
 that describes the area it can monitor. 
 Each drone can only communicate directly with its own neighbourhood $\mathcal{N}$, 
 which in this case depends on a $\mathbb{O}$ range,
From the neighbourhood, 
 each node can also perceive the direction of the other nodes.
%
Each node moves following a certain action composed of two components ($\alpha$, $\theta$) 
 which respectively describe the angle and intensity of the movement (i.e., the velocity vector).
\subsection{Goal}
 The objective of this scenario is to:
\begin{itemize}
\item Maximise the number of drones within the phenomena
\item Minimise the number of drones without a neighbourhood
\item Maximise the coverage of the system
\end{itemize}

@@

 Finally, to maximise the coverage, 
  we decided to calculate the maximum and minimum distance of the neighbourhood and give more value in case these two values are similar. 
%
This means that the average distance will tend towards the one expressed by the viewing range of each node:
% 
\begin{equation*}
R^{C}_{v} = \frac{d_{max} - d_{min}}{\mathbb{O}}
\end{equation*}
The final reward function is defined as:
\begin{equation*}
R_{v} = R^a_{v} + R^{\mathcal{N}}_{v} + R^{C}_{v}
\end{equation*}
